---
title: "Problem Set 4"
author: "Larry Schwerzler"
date: "November 22, 2015"
output: html_document
---

# 4.1 price vs. x
In this problem set, you'll continue to explore the diamonds data set.

Your first task is to create a scatterplot of price vs x. using the ggplot syntax.

```{r}
data("diamonds")
ggplot(aes(x, price), data = diamonds) + geom_point()
```

# 4.2 Findings - price vs. x
What are your observations about the scatterplot of price vs x?

There seems to be a exponential relationship between the two increasing x leads to an exponential increase in price. 
Most of the data is grouped between x's of 4 and 10

# 4.3 Correlations
What is the Correlation between price and x?
```{r}
cor.test(diamonds$price, diamonds$x)
```
0.8844352

What is the Correlation between price and y?
```{r}
cor.test(diamonds$price, diamonds$y)
```
0.8654209

What is the Correlation between price and z?
```{r}
cor.test(diamonds$price, diamonds$z)
```
0.8612494

# 4.4  price vs. depth
Create a simple scatter plot of price vs depth.
```{r}
ggplot(aes(price, depth), data=diamonds) + geom_point()
```

# 4.5 Adjustments - price vs. depth
Change the code to make the transparency of the points to be 1/100 of what they are now and mark the x-axis every 2 units. See the instructor notes for two hints.
```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha = 1/100) + 
  scale_x_continuous(breaks = seq(43,79,2))
```

# 4.6 Typical Depth Range
Based on the scatterplot of depth vs price, most of the diamonds are between what values of depth?
59,64

# 4.7 Correlation - price and depth
What is the correlation of depth vs price?
```{r}
cor.test(diamonds$depth, diamonds$price)
```
-0.0106474 

Based on the correlation coefficient would you use depth to predict the price of diamonds? 
No.

Why?
There is almost no correlation between the two. Meaning a change in one is not predictive of a change in the other. 

# 4.8 price vs. carat
Create a scatterplot of price vs carat and omit the top 1% of price and carat values.
```{r}
ggplot(aes(carat,price), data = diamonds) + geom_point() + ylim(0, quantile(diamonds$price, .9)) + xlim(0,quantile(diamonds$carat,.9))
```

# 4.9 price vs. volume
Create a scatterplot of price vs. volume (x * y * z). This is a very rough approximation for a diamond's volume.

Create a new variable for volume in the diamonds data frame. This will be useful in a later exercise.
```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ggplot(aes(volume, price), data = diamonds) + geom_point()
```

# 4.10 Findings - price vs. volume
What are your observations from the price vs volume scatter plot?

as volume goes up price increases exponentially. 
There are a few outliers including one diamond that appears to have a very large volume, without a comparatively large price. 

# 4.11 Correlations on Subsets
What is the correlation between price and volume? Exclude diamonds that have a volume of 0 or that are greater than or equal to 800. 
```{r}
diamonds.volumes_subset <- subset(diamonds, volume <= 800 & volume > 0)
cor.test(diamonds.volumes_subset$volume, diamonds.volumes_subset$price)
```
0.9235455

# 4.12 Adjustments - price vs. volume
Subset the data to exclude diamonds with a volume greater than or equal to 800. Also, exclude diamonds with a volume of 0. Adjust the transparency of the points and add a linear model to the plot. (See the Instructor Notes or look up the documentation of geom_smooth() for more details about smoothers.)

We encourage you to think about this next question and to post your thoughts in the discussion section.

Do you think this would be a useful model to estimate the price of diamonds? Why or why not?
```{r}
diamonds.volumes_subset <- subset(diamonds, volume < 800 & volume > 0)
ggplot(aes(volume, price), data = diamonds.volumes_subset) + geom_point(alpha = 1/20) + geom_smooth(model = 'lm')
```

# 4.13 Mean Price by Clarity
Use the function dplyr package to create a new data frame containing info on diamonds by clarity.

Name the data frame diamondsByClarity

The data frame should contain the following variables in this order.

     (1) mean_price
     (2) median_price
     (3) min_price
     (4) max_price
     (5) n

where n is the number of diamonds in each level of clarity.
```{r}
diamondsByClarity <- diamonds %>% group_by(clarity) %>%
  summarise(mean_price = mean(price),
            median_price = median(as.numeric(price)),
            min_price = min(price),
            max_price = max(price),
            n = n())
```

# 4.14 Bar Charts of Mean Price
We've created summary data frames with the mean price by clarity and color. You can run the code in R to verify what data is in the variables diamonds_mp_by_clarity and diamonds_mp_by_color.

Your task is to write additional code to create two bar plots on one output image using the grid.arrange() function from the package gridExtra.
```{r}
data(diamonds)
library(dplyr)

diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

p1 = ggplot(aes(x = clarity, y = mean_price), data = diamonds_mp_by_clarity) + geom_bar(stat= "identity")
p2 = ggplot(aes(x = color, y = mean_price), data = diamonds_mp_by_color) + geom_bar(stat= "identity")
library(gridExtra)
grid.arrange(p1,p2, ncol = 1)
```

# 4.15 Trends in Mean Price
What did you notice in each of the bar charts for mean price by clarity and mean price by color. 

Clarity has a peak and a trailing off of mean price as you move from the lesser clarity grades to the higher clarity grades. 

Color mean price increases as the color gets better

# 4.16 Gapminder Revisited
The Gapminder website contains over 500 data sets with information about the world's population. Your task is to continue the investigation you did at the end of Problem Set 3 or you can start fresh and choose a different data set from Gapminder.

If you're feeling adventurous or want to try some data munging see if you can find a data set or scrape one from the web.

In your investigation, examine pairs of variable and create 2-5 plots that make use of the techniques from Lesson 4.
```{r}
library(tidyr)
library(gridExtra)
cp100 <- read.csv('cell phone per 100.csv', sep=',', stringsAsFactors=FALSE, check.names = FALSE)
cp100.reshaped <- cp100 %>% gather(year, cp100, c(2:48))
cp100.reshaped$year <- as.numeric(as.character(cp100.reshaped$year))
cp100.grouped_by_year <- cp100.reshaped %>% group_by(year) %>%
  summarise(cp100mean = mean(cp100,na.rm=TRUE),
            n = n()) %>%
  arrange(year)
  
p1 <- ggplot(aes(year, cp100mean), data = subset(cp100.grouped_by_year, year >= 1985)) + geom_bar(stat="identity", fun.y = mean) + labs(title = "Global Cellphones per 100 people", x = 'Year', y = 'Cellphones Per 100 People') 

p2 <- ggplot(aes(year, cp100), data = subset(cp100.reshaped, year >= 1985 & Country == "Japan")) + geom_bar(stat="identity")+ labs(title = "Japanese Cellphones per 100 people", x = 'Year', y = 'Cellphones Per 100 People')
p3 <- ggplot(aes(year, cp100), data = subset(cp100.reshaped, year >= 1985 & Country == "United States")) + geom_bar(stat="identity") + labs(title = "US Cellphones per 100 people", x = 'Year', y = 'Cellphones Per 100 People')

grid.arrange(p1,p2,p3, ncol=1)
```
