---
title: "Problem Set 5"
author: "Larry Schwerzler"
date: "November 22, 2015"
output: html_document
---
# 5.1 Price Histograms with Facet and Color
Create a histogram of diamond prices. Facet the histogram by diamond color and use cut to color the histogram bars.
```{r}
data("diamonds")
ggplot(aes(x= price), data = diamonds) + geom_histogram(aes(color = cut)) + facet_wrap(~color) + scale_color_brewer(type = 'qual')
```

# 5.2 Price vs. Table Colored by Cut
Create a scatterplot of diamond price vs. table and color the points by the cut of the diamond.

The plot should look something like this. http://i.imgur.com/rQF9jQr.jpg

Note: In the link, a color palette of type 'qual' was used to color the scatterplot using scale_color_brewer(type = 'qual')

```{r}
ggplot(aes(x=table, y=price), data=diamonds) + geom_point(aes(color=cut)) + scale_color_brewer(type='qual')
```

# 5.3 Typical Table Value
what is the typical table range for the majority of diamonds of ideal cut?
54 to 57
What is the typical table range for the majority of diamonds of premium cut?
58 to 62
use the graph you created from the previous excercise you do not need to run summaries. 

# 5.4 Price vs. Volume and Diamond Clarity
Create a scatterplot of diamond price vs. volume (x * y * z) and color the points by the clarity of diamonds. Use scale on the y-axis to take the log10 of price. You should also omit the top 1% of diamond volumes from the plot.

Note: Volume is a very rough approximation of a diamond's actual volume.

The plot should look something like this. http://i.imgur.com/excUpea.jpg

Note: In the link, a color palette of type 'div' was used to color the scatterplot using scale_color_brewer(type = 'div')
```{r}
diamonds <- transform(diamonds, volume = x * y *z)
ggplot(aes(x=volume, y= price), data = subset(diamonds, volume < quantile(diamonds$volume, .9))) + geom_point(aes(color=clarity)) + scale_color_brewer(type='div') + scale_y_log10()
```

# 5.5 Proportion of Friendships Initiated
Many interesting variables are derived from two or more others. For example, we might wonder how much of a person's network on a service like Facebook the user actively initiated. Two users with the same degree (or number of friends) might be very different if one initiated most of those connections on the service, while the other initiated very few. So it could be useful to consider this proportion of existing friendships that the user initiated. This might be a good predictor of how active a user is compared with their peers, or other traits, such as personality (i.e., is this person an extrovert?).

Your task is to create a new variable called 'prop_initiated' in the Pseudo-Facebook data set. The variable should contain the proportion of friendships that the user initiated.
```{r}
pf <- read.delim('pseudo_facebook.tsv')
pf <- transform(pf, prop_initiated = friendships_initiated/friend_count)
```

# 5.6 prop_initiated vs. tenure
Create a line graph of the median proportion of friendships initiated ('prop_initiated') vs. tenure and color the line segment by year_joined.bucket.

Recall, we created year_joined.bucket in Lesson 5 by first creating year_joined from the variable tenure. Then, we used the cut function on year_joined to create four bins or cohorts of users.

(2004, 2009]
(2009, 2011]
(2011, 2012]
(2012, 2014]

The plot should look something like this. http://i.imgur.com/vNjPtDh.jpg OR this http://i.imgur.com/IBN1ufQ.jpg

```{r}
pf$year_joined <- floor(2014 - (pf$tenure/365))
pf$year_joined.bucket <- cut(pf$year_joined, breaks = c(2004,2009,2011,2012,2014))
ggplot(aes(x=tenure, y= prop_initiated), data = pf) + geom_line(aes(color=year_joined.bucket), stat="summary",fun.y = median)
```

# 5.7 Smoothing prop_initiated vs. tenure
Smooth the last plot you created of of prop_initiated vs tenure colored by year_joined.bucket. You can bin together ranges of tenure or add a smoother to the plot.

There won't be a solution image for this exercise. You will answer some questions about your plot in the next two exercises.

```{r}
ggplot(aes(x=tenure, y= prop_initiated), data = pf) + geom_smooth(aes(color=year_joined.bucket))
```

# 5.8 Greatest prop_initiated Group
On average which group initated the greatest proportion of its facebook friendships? The plot with the smoother that you created in the last excercise can help you answer this question?
People who joined after 2012

# 5.9 Largest Group Mean prop_initiated
For the group with the largest proportion of friendships initated, what is the groups average (mean) proportion of friendships initiated?
```{r}
aggregate(pf$prop_initiated,by=list(pf$year_joined.bucket), mean)
```

# 5.10 Price/Carat Binned, Faceted, & Colored
Create a scatter plot of the price/carat ratio of diamonds. The variable x should be assigned to cut. The points should be colored by diamond color, and the plot should be faceted by clarity.

The plot should look something like this. http://i.imgur.com/YzbWkHT.jpg.

Note: In the link, a color palette of type 'div' was used to color the histogram using scale_color_brewer(type = 'div')
```{r}
ggplot(aes(x=cut, y = price/carat), data = diamonds)+ geom_jitter(aes(color=color)) + facet_wrap(~clarity) + scale_color_brewer(type = 'div')
```

# 5.11 Gapminder Multivariate Analysis
The Gapminder website contains over 500 data sets with information about the world's population. Your task is to continue the investigation you did at the end of Problem Set 4 or you can start fresh and choose a different data set from Gapminder.

If you're feeling adventurous or want to try some data munging see if you can find a data set or scrape one from the web.

In your investigation, examine 3 or more variables and create 2-5 plots that make use of the techniques from Lesson 5.
```{r}
library(tidyr)
set.seed(6813)
bmi <- read.csv('BMI.csv',sep=',', stringsAsFactors=FALSE, check.names = FALSE)
bmi.reshaped <- bmi %>% gather(year, bmi, c(3:31))
bmi.reshaped$year <- as.numeric(as.character(bmi.reshaped$year))
sample.countries <- sample(bmi.reshaped$Country, 16)
ggplot(aes(x=year, y= bmi), data = subset(bmi.reshaped, Country %in% sample.countries )) + geom_line(aes(color=Gender), stat="summary", fun.y = mean)  + facet_wrap(~Country)
```
